Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-06 ms.tex
random fixes

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius).  In the case of sufficiently weak heating,
however, no stagnation radius may exist at all, i.e. the entire ISM of
the galaxy is feeding the SMBH.  Another goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our numerical results.  In $\S\ref{sec:discussion}$ we discuss the implications of our results, to topics including...   In $\S\ref{sec:conclusions}$ we summarize our conclusions.  Table \ref{table:definitions} provides the definitions of commonly used variables.  Appendix \ref{app:rs} provides useful analytic results, while Appendix \ref{app:windheat} describes our method for calculating stellar wind heating and mass input in detail.

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{lc}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating rate from stellar velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from stellar velocity dispersion \\
$r_{s}$ & Stagnation radius (where radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\  
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
The use of equation (\ref{eq:Msigma}) is questionable for low mass black holes (e.g., \citealt{Greene&Ho07}), but our results are not overly sensitive to this assumption.

@@

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\be
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\ee
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}; $\S\ref{sec:instability}$) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:cooling}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the steady-state inflow/outflow solution (assuming one exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.  Also note that our solutions can be scaled to any value of the mass input parameter, $\eta$, since the mass and energy source terms scale linearly with $\rho$ or $\rho_{\star}$; however, the precise value of $\eta$ must be specified when cooling or thermal conduction are included, since these scale as gas density {\it squared}.

We assess the accuracy of our numerical solutions by checking that mass is conserved across the grid, in addition to the integral constraint on the energy (Bernoulli integral).  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this paper.  {\bf BDM:
  more details...}

@@

To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n kT/2)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the stagnation radius, as can be derived using the equality 
\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations
\eqref{eq:mdot_analytic},\eqref{eq:mdot_gas}, and
\eqref{eq:rhostarrs}.  \citet{McCourt+12} argue cooling instability
develops in a hydrostatic atmosphere if $t_{\rm cool} \ll t_{\rm ff}$,
so equation (\ref{eq:cooling2}) represents a good proxy for
instability in this case as well.

@@

  \subsection{Angular Momentum}
  \label{sec:ang}

@@

\citet{EmsellemCappellari+:2007a} (their Fig.~2) find that $\lambda_R$ is generally $< 0.1$ on radial scales less that 10 per cent of the galaxy half-light radius and that $\lambda_R$ is a decreasing function of $R$.  From equation (\ref{eq:rcirc}) we conclude that $r_{\rm circ} \lesssim 0.01r_{s}$, a location well inside the inner sonic point of the flow (typically located at $\sim 0.1r_{\rm s}$; Fig.~\ref{fig:profiles}).  This suggests that angular momentum is only becomes important for thermally stable flows in a region that is causally disconnected from the rest of the flow.  To first order we are thus justifying in neglecting its effects on our quantities such as the mass accretion rate.

>>>>

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius).  In the case of sufficiently weak heating,
however, no stagnation radius may exist at all, i.e. in principle the entire ISM of
the galaxy is feeding the SMBH.  Another goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our numerical results.  In $\S\ref{sec:discussion}$ we discuss the implications of our results, to topics including...   In $\S\ref{sec:conclusions}$ we summarize our conclusions.  Table \ref{table:definitions} provides the definitions of commonly used variables.  Appendix \ref{app:rs} provides useful analytic results, while Appendix \ref{app:windheat} describes our method for calculating stellar wind heating and mass input in greater detail.

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{lc}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating rate from stellar velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from stellar velocity dispersion \\
$r_{s}$ & Stagnation radius (where radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\ 
$r_{\rm Ia}$ & radius exterior to which Ia SN are infrequent compared to the dynamical/inflow timescale (eq.~[\ref{eq:rIa}]) \\ 
$\rho_{\star}(r)$ & 3D radial stellar density profile \\
$\rho(r)$ & gas density of CNM \\
$M_{\star}(r)$ & total enclosed stellar mass inside radius $r$ \\
$M_{\rm enc}(r)$ & total encolsed mass inside radius $r$ (SMBH + stars) \\
$q(r)$ & mass source term due to stellar winds (eq.~[\ref{eq:q}]) \\
$\eta$ & parameter setting normalization of mass input from stellar winds (eq.~[\ref{eq:q}]) \\
$\tau_{\star}$ & age of stellar population, in case of a single burst of star formation \\
$\Gamma$ & power-law slope of radial stellar density profile interior to the break radius \\
$\sigma(r)$ & stellar velocity dispersion \\
$\dot{M}$ & mass accretion rate \\
$\dot{M}_{\rm Ia}$ & maximum mass accretion rate as limited by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]) \\
$\dot{M}_{\rm C}$ & equilibrium accretion rate set by Compton heating (eq.~[\ref{eq:MdotC}]) \\
$\dot{M}_{\rm TI}$ & maximum accretion rate for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]) \\
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ & ratio of external heating (stellar winds, SN Ia, MSPs) to radiative cooling (eq.~[\ref{eq:cooling2}]) \\
$t_{\rm h}$ & Hubble timescale \\
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta > \Gamma$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of a better alternative.

@@

Radiative cooling contributes an additional term to equation (\ref{eq:dsdt}) of the form
\be
\dot{q}_{\rm rad} = -\Lambda(T)n^{2},
\label{eq:qdot_rad}
\ee
where $n = \rho/\mu m_p$ is the gas number density and $\Lambda(T)$ is the cooling function.  We neglect radiative cooling in our baseline calculations, despite the fact that this is not justified when the wind heating $\vwO$ is low or if the mass return rate $\eta$ is high.  However, once radiative cooling becomes important compared to other sources of heating and cooling, its presence leads to thermal instability (e.g.~\citealt{McCourt+12}; $\S\ref{sec:instability}$) that cannot be accounted for by our 1D
time-independent model.  A main goal of this work is to use solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities are most likely to develop ($\S\ref{sec:instability}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the steady-state inflow/outflow solution (assuming one exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.  Also note that our solutions can be scaled to any value of the mass input parameter, $\eta$, since the mass and energy source terms scale linearly with $\rho$ or $\rho_{\star}$; however, the precise value of $\eta$ must be specified when cooling or thermal conduction are included, since these scale as gas density {\it squared}.  We check the accuracy of our numerical solutions by confirming that mass is conserved across the grid, in addition to the integral constraint on the energy (Bernoulli integral).

@@

To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n kT/2\mu )/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the stagnation radius, as can be derived using the equality 
\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations
\eqref{eq:mdot_analytic},\eqref{eq:mdot_gas}, and
\eqref{eq:rhostarrs}.  \citet{McCourt+12} argue cooling instability
develops in a hydrostatic atmosphere if $t_{\rm cool} \ll t_{\rm ff}$,
so equation (\ref{eq:cooling2}) represents a good proxy for
instability in this case as well.

@@

  \subsection{Angular Momentum}
  \label{sec:rotation}

@@

\citet{EmsellemCappellari+:2007a} (their Fig.~2) find that $\lambda_R$ is generally $< 0.1$ on radial scales less that 10 per cent of the galaxy half-light radius and that $\lambda_R$ is a decreasing function of $R$.  From equation (\ref{eq:rcirc}) we conclude that $r_{\rm circ} \lesssim 0.01r_{s}$, a location well inside the inner sonic point of the flow (typically located at $\sim 0.1r_{\rm s}$; Fig.~\ref{fig:profiles}).  This suggests that angular momentum is only becomes important for thermally stable flows in a region that is causally disconnected from the rest of the flow.  We are thus justifying in neglecting the effects of rotation in calculating the mass accretion rate.
